---
title: "Differential Expression Analysis with EdgeR"
author: "Carson Stacy & Jeffrey Lewis"
date: "Fall 2023"
output:
  html_document:
    code_folding: show
editor_options:
  markdown:
    wrap: 72
---

last updated: `r Sys.Date()`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed("1492")
```

# Getting Things Setup

As usual, make sure we have the right packages for this exercise

```{r ready-packages}
if (!require("pacman")) install.packages("pacman"); library(pacman)

# let's load all of the files we were using and want to have again today
p_load("tidyverse", "knitr", "readr",
       "pander", "BiocManager", 
       "dplyr", "stringr", 
       "statmod", # required dependency, need to load manually on some macOS versions.
       "purrr", # for working with lists (beautify column names)
       "webshot2", # allow for pdf of output table.
       "reactable") # for pretty tables.

# We also need these Bioconductor packages today.
p_load("edgeR", "AnnotationDbi", "org.Sc.sgd.db")
```

# Description

This will be our first differential expression analysis workflow,
converting gene counts across samples into meaningful information about
genes that appear to be significantly differentially expressed between
samples

# Learning outcomes

At the end of this exercise, you should be able to:

-   Generate a table of sample metadata.
-   Filter low counts and normalize count data.
-   Utilize the edgeR package to identify differentially expressed
    genes.

```{r load-libraries}
library(edgeR)
library(org.Sc.sgd.db)
# for ease of use, set max number of digits after decimal
options(digits=3)
```

# Loading in the featureCounts object

We saved this file in the last exercise (0Read_Counting.Rmd) from the
RSubread package. Now we can load that object back in and assign it to
the variable `fc`. Be sure to change the file path if you have saved it
in a different location.

```{r load-count_subset, error=TRUE}
path_fc_object <- path.expand("~/Desktop/Genomic_Data_Analysis/Data/Counts/Rsubread/rsubread.yeast_fc_output.Rds")

counts_subset <- readRDS(file = path_fc_object)$counts
```

We generated those counts on a subset of the fastq files, but we can
load the complete count file with the command below. This file has been
generated with the full size fastq files with Salmon.

```{r load-full_counts}
counts <- read_tsv('https://github.com/clstacy/GenomicDataAnalysis_Fa23/raw/main/data/ethanol_stress/counts/salmon.gene_counts.merged.nonsubsamp.tsv',
                   col_names = TRUE) %>%
  # when we saved the tsv file, it converted the rownames to a column,
  # we are converting it back with this piped command.
  column_to_rownames("Name")
```

So far, we've been able to process all of the fastq files without much
information about what each sample is in the experimental design. Now,
we need the metadata for the samples. Note that the order matters for
these files

To find the order of files we need, we can get just the part of the
column name before the first "." symbol with this command:

```{r identify-sampleOrder}
str_split_fixed(counts %>% colnames(), "\\.", n = 2)[, 1] %>% cat()
```

```{r generate-metadata}
sample_metadata <- tribble(
  ~Sample,                      ~Genotype,    ~Condition,
  "YPS606_MSN24_ETOH_REP1_R1",   "msn24dd",   "EtOH",
  "YPS606_MSN24_ETOH_REP2_R1",   "msn24dd",   "EtOH",
  "YPS606_MSN24_ETOH_REP3_R1",   "msn24dd",   "EtOH",
  "YPS606_MSN24_ETOH_REP4_R1",   "msn24dd",   "EtOH",
  "YPS606_MSN24_MOCK_REP1_R1",   "msn24dd",   "unstressed",
  "YPS606_MSN24_MOCK_REP2_R1",   "msn24dd",   "unstressed",
  "YPS606_MSN24_MOCK_REP3_R1",   "msn24dd",   "unstressed",
  "YPS606_MSN24_MOCK_REP4_R1",   "msn24dd",   "unstressed",
  "YPS606_WT_ETOH_REP1_R1",      "WT",        "EtOH",
  "YPS606_WT_ETOH_REP2_R1",      "WT",        "EtOH",
  "YPS606_WT_ETOH_REP3_R1",      "WT",        "EtOH",
  "YPS606_WT_ETOH_REP4_R1",      "WT",        "EtOH",
  "YPS606_WT_MOCK_REP1_R1",      "WT",        "unstressed",
  "YPS606_WT_MOCK_REP2_R1",      "WT",        "unstressed",
  "YPS606_WT_MOCK_REP3_R1",      "WT",        "unstressed",
  "YPS606_WT_MOCK_REP4_R1",      "WT",        "unstressed") %>%
  # Create a new column that combines the Genotype and Condition value
  mutate(Group = factor(
    paste(Genotype, Condition, sep = "."),
    levels = c(
      "WT.unstressed","WT.EtOH",
      "msn24dd.unstressed", "msn24dd.EtOH"
    )
  )) %>%
  # make Condition and Genotype a factor (with baseline as first level) for edgeR
  mutate(
    Genotype = factor(Genotype,
                      levels = c("WT", "msn24dd")),
    Condition = factor(Condition,
                       levels = c("unstressed", "EtOH"))
  )
  

```

Now, let's create a design matrix with this information

```{r}
group <- sample_metadata$Group
design <- model.matrix(~ 0 + group)
design
colnames(design) <- levels(group)
design
```

# Count loading and Annotation

The count matrix is used to construct a DGEList class object. This is
the main data class in the edgeR package. The DGEList object is used to
store all the information required to fit a generalized linear model to
the data, including library sizes and dispersion estimates as well as
counts for each gene.

```{r load-counts}
y <- DGEList(counts, group=group)
colnames(y) <- sample_metadata$Sample
y$samples
```

Human-readable gene symbols can also be added to complement the gene ID
for each gene, using the annotation in the org.Sc.sgd.db package.

```{r}
y$genes <- AnnotationDbi::select(org.Sc.sgd.db,keys=rownames(y),columns="GENENAME")

head(y$genes)
```

# Filtering to remove low counts

Genes with very low counts across all libraries provide little evidence
for differential expression. In addition, the pronounced discreteness
of these counts interferes with some of the statistical approximations
that are used later in the pipeline. These genes should be filtered out
prior to further analysis. Here, we will retain a gene only if it is
expressed at a count-per-million (CPM) above 0.7 in at least four
samples.

```{r filter-lowCounts}
keep <- rowSums(cpm(y) > 0.7) >= 4
y <- y[keep,]
summary(keep)
```

Where did those cutoff numbers come from?

As a general rule, we don't want to exclude a gene that is expressed in
only one group, so a cutoff number equal to the number of replicates can
be a good starting point. For counts, a good threshold can be chosen by
identifying the CPM that corresponds to a count of 10, which in this
case would be about 0.7:

```{r estimate-cpmCutoff}
cpm(10, mean(y$samples$lib.size))
```

Smaller CPM thresholds are usually appropriate for larger libraries.

# Normalization for composition bias

TMM normalization is performed to eliminate composition biases between
libraries. This generates a set of normalization factors, where the
product of these factors and the library sizes defines the effective
library size. The calcNormFactors function returns the DGEList argument
with only the norm.factors changed.

```{r calc-normFactors}
y <- calcNormFactors(y)
y$samples
```

The normalization factors multiply to unity across all libraries. A
normalization factor below unity indicates that the library size will be
scaled down, as there is more suppression (i.e., composition bias) in
that library relative to the other libraries. This is also equivalent to
scaling the counts upwards in that sample. Conversely, a factor above
unity scales up the library size and is equivalent to downscaling the
counts. The performance of the TMM normalization procedure can be
examined using mean- difference (MD) plots. This visualizes the library
size-adjusted log-fold change between two libraries (the difference)
against the average log-expression across those libraries (the mean).
The below command plots an MD plot, comparing sample 1 against an
artificial library constructed from the average of all other samples.

## MDS plots

```{r plotMDS, fig.show="hold", out.width="25%"}
for (sample in 1:nrow(y$samples)) {
  plotMD(cpm(y, log=TRUE), column=sample)
  abline(h=0, col="red", lty=2, lwd=2)
}
```

# Exploring differences between libraries

The data can be explored by generating multi-dimensional scaling (MDS)
plots. This visualizes the differences between the expression profiles
of different samples in two dimensions. The next plot shows the MDS plot
for the yeast heatshock data.

```{r plot-MDS}
points <- c(1,1,2,2)
colors <- rep(c("black", "red"),8)
plotMDS(y, col=colors[group], pch=points[group])
legend("topright", legend=levels(group),
     pch=points, col=colors, ncol=2)
title(main="PCA plot")
```

# Estimate Dispersion

The trended NB dispersion is estimated using the estimateDisp function.
This returns the DGEList object with additional entries for the
estimated NB dispersions for all genes. These estimates can be
visualized with plotBCV, which shows the root-estimate, i.e., the
biological coefficient of variation for each gene

```{r estimate-dispersion}
y <- estimateDisp(y, design, robust=TRUE)
plotBCV(y)
title(main="Biological Coefficient of Variation (BCV) vs gene abundance")
```

In general, the trend in the NB dispersions should decrease smoothly
with increasing abundance. This is because the expression of
high-abundance genes is expected to be more stable than that of
low-abundance genes. Any substantial increase at high abundances may be
indicative of batch effects or trended biases. The value of the trended
NB dispersions should range between 0.005 to 0.05 for
laboratory-controlled biological systems like mice or cell lines, though
larger values will be observed for patient-derived data (\> 0.1)

For the QL dispersions, estimation can be performed using the glmQLFit
function. This returns a DGEGLM object containing the estimated values
of the GLM coefficients for each gene

```{r generate-fit}
fit <- glmQLFit(y, design, robust=TRUE)
head(fit$coefficients)
plotQLDisp(fit)
title(main="QL Dispersion of the fit")
```

EB squeezing of the raw dispersion estimators towards the trend reduces
the uncertainty of the final estimators. The extent of this moderation
is determined by the value of the prior df, as estimated from the data.
Large estimates for the prior df indicate that the QL dispersions are
less variable between genes, meaning that stronger EB moderation can be
performed. Small values for the prior df indicate that the dispersions
are highly variable, meaning that strong moderation would be
inappropriate

Setting `robust=TRUE` in glmQLFit is strongly recommended. This causes
glmQLFit to estimate a vector of df.prior values, with lower values for
outlier genes and larger values for the main body of genes.

# Testing for differential expression

The final step is to actually test for significant differential
expression in each gene, using the QL F-test. The contrast of interest
can be specified using the makeContrasts function. Here, genes are
detected that are DE between the stressed and unstressed. This is done
by defining the null hypothesis as heat stressed - unstressed = 0.

```{r fit-contrast}
# generate contrasts we are interested in learning about
my.contrasts <- makeContrasts(EtOHvsMOCK.WT = WT.EtOH - WT.unstressed, 
                     EtOHvsMOCK.MSN24dd = msn24dd.EtOH - msn24dd.unstressed,
                     EtOH.MSN24ddvsWT = msn24dd.EtOH - WT.EtOH,
                     MOCK.MSN24ddvsWT = msn24dd.unstressed - WT.unstressed,
                     EtOHvsWT.MSN24ddvsWT = (msn24dd.EtOH-msn24dd.unstressed)-(WT.EtOH-WT.unstressed),
                     levels=design)

# This contrast looks at the difference in the stress responses between mutant and WT
res <- glmQLFTest(fit, contrast = my.contrasts[,"EtOHvsWT.MSN24ddvsWT"])
```

```{r create-Table}
# let's take a quick look at the results
topTags(res, n=10) 

# generate a beautiful table for the pdf/html file.
topTags(res, n=Inf) %>% data.frame() %>% 
  arrange(FDR) %>%
  mutate(logFC=round(logFC,2)) %>%
  # mutate(across(where(is.numeric), signif, 3)) %>%
  mutate_if(is.numeric, signif, 3) %>%
  remove_rownames() %>%
  reactable(
    searchable = TRUE,
    showSortable = TRUE,
    columns = list(ORF = colDef(
      cell = function(value) {
        # Render as a link
        url <-
          sprintf("https://www.yeastgenome.org/locus/%s", value)
        htmltools::tags$a(href = url, target = "_blank", as.character(value))
      }
    ))
  )
```

```{r summarize-DEgenes}
is.de <- decideTestsDGE(res, 
                        p.value=0.05,
                        lfc = 0) # this allows you to set a cutoff, BUT...
# if you want to compare against a FC that isn't 0, should use glmTreat instead.

summary(is.de)
```

Let's take a quick look at the differential expression

```{r visualize-DEgenes}
plotSmear(res, de.tags=rownames(res)[is.de!=0])
title(main="DE genes using glmQLFTest, FDR<0.05")
```

Here is how we can save our output file(s).

```{r save-res}
# Choose topTags destination
dir_output_edgeR <-
  path.expand("~/Desktop/Genomic_Data_Analysis/Analysis/edgeR/")
if (!dir.exists(dir_output_edgeR)) {
  dir.create(dir_output_edgeR, recursive = TRUE)
}

# for shairng with others, the topTags output is convenient.
topTags(res, n = Inf) %>% data.frame() %>%
  arrange(desc(logFC)) %>%
  mutate(logFC = round(logFC, 2)) %>%
  # mutate(across(where(is.numeric), signif, 3)) %>%
  mutate_if(is.numeric, signif, 3) %>%
  write_tsv(., file = paste0(dir_output_edgeR, "yeast_topTags_edgeR.tsv"))

# for subsequent analysis, let's save the res object as an R data object.
saveRDS(object = res, file = paste0(dir_output_edgeR, "yeast_res_edgeR.Rds"))

# we might also want our y object list
saveRDS(object = y, file = paste0(dir_output_edgeR, "yeast_y_edgeR.Rds"))
```

# Looking at all contrasts at once

If we want results from all contrasts, we need to loop through them in edgeR, and them combine the results We will look more at the results of this in a later activity.
```{r}
# One way is to not specify just one contrast, like this:
res_all <- glmQLFTest(fit, contrast = my.contrasts)

res_all %>% 
  topTags(n=Inf) %>% 
  data.frame()
  head()

# alternatively, we can loop to get DE genes in each contrast.
# here we are just saving which genes are DE per contrast
decideTests_edgeR_tmp <- list()
for (i in 1:ncol(my.contrasts)){

    current.res <- glmQLFTest(fit, contrast = my.contrasts[,paste0(dimnames(my.contrasts)$Contrasts[i])])
    # current.res <- eBayes(current.res)
    decideTests_edgeR_tmp[[i]] <- current.res %>% decideTests(p.value = 0.05, lfc = 0) %>%
  as.data.frame()

}

decideTests_edgeR <- list_cbind(decideTests_edgeR_tmp) %>%
  rownames_to_column("gene")

head(decideTests_edgeR)

# save this file for future analysis
write_tsv(decideTests_edgeR, "~/Documents/GitHub/GenomicDataAnalysis_Fa23/analysis/yeast_decideTests_allContrasts_edgeR.tsv")

# for subsequent analysis, let's also save the res_all object as an R data object.
saveRDS(object = res_all, file = paste0(dir_output_edgeR, "yeast_res_all_edgeR.Rds"))
```


## Questions

Question 1: How many genes were upregulated and downregulated in the
contrast we looked at in todays activity? Be sure to clarify the cutoffs
used for determining significance.

Question 2: Which gene has the lowest pvalue with a postive log2 fold
change?

Question 3: Choose one of the contrasts in `my.contrasts` that we didn't
test together, and identify the top 3 most differentially expressed
genes.

Question 4: In the contrast you chose, give a brief description of the
biological interpretation of that contrast.

Question 5: In the example above, we tested for differential expression of any magnitude. Often, we only care about changes of at least a certain magnitude. In this case, we need to use a different command. using the same data, test for genes with differential expression of at least 1 log2 fold change using the `glmTreat` function in edgeR. How do these results compare to DE genes without a logFC cutoff?

## A template set of code chunks for doing this is below:

We already loaded in the salmon counts as the object `counts`
above. This code chunk just re-downloads that same file.

```{r load-data}
path_salmon_counts <- 'https://github.com/clstacy/GenomicDataAnalysis_Fa23/raw/main/data/ethanol_stress/counts/salmon.gene_counts.merged.nonsubsamp.tsv'

counts <- read_tsv(path_salmon_counts,
                   col_names = TRUE) %>%
  column_to_rownames("Name")
```

```{r repeat-edgeRWorkflow}
# We are reusing the sample_metadata, group, etc that we assigned above

# create DGEList with salmon counts
y <- DGEList(counts, group=group)
colnames(y) <- sample_metadata$Sample

# add gene names
y$genes <- AnnotationDbi::select(org.Sc.sgd.db,keys=rownames(y),
                                        columns="GENENAME")

# filter low counts
keep <- rowSums(cpm(y) > 60) >= 4
y <- y[keep,]

# calculate norm factors
y <-  calcNormFactors(y)

# estimate dispersion
y <- estimateDisp(y, design, robust=TRUE)

# generate the fit
fit <- glmQLFit(y, design, robust=TRUE)


# Note that, unlike other edgeR functions such as glmLRT and glmQLFTest, 
# glmTreat can only accept a single contrast. 
# If contrast is a matrix with multiple columns, then only the first column will be used.

# Implement a test against FC at least 1 the test our contrast of interest
tr <- glmTreat(fit, 
               contrast = my.contrasts[,"EtOHvsWT.MSN24ddvsWT"],
               lfc=1)

# generate a beautiful table for the pdf/html file.
topTags(tr, n = Inf) %>%
  data.frame() %>%
  arrange(FDR) %>%
  mutate(logFC = round(logFC, 2)) %>%
  # mutate(across(where(is.numeric), signif, 3)) %>%
  mutate_if(is.numeric, signif, 3) %>%
  remove_rownames() %>%
  reactable(
    searchable = TRUE,
    showSortable = TRUE,
    columns = list(ORF = colDef(
      cell = function(value) {
        # Render as a link
        url <-
          sprintf("https://www.yeastgenome.org/locus/%s", value)
        htmltools::tags$a(href = url, target = "_blank", as.character(value))
      }
    ))
  )

# write the table to a tsv file
topTags(tr, n=Inf) %>% 
  data.frame() %>% 
  arrange(FDR) %>%
  mutate(logFC=round(logFC,2)) %>%
  # mutate(across(where(is.numeric), signif, 3)) %>%
  mutate_if(is.numeric, signif, 3) %>%
  write_tsv(., file = paste0(dir_output_edgeR, "yeast_lfc1topTags_edgeR.tsv"))

# summarize the DE genes
is.de_tr <- decideTestsDGE(tr, p.value=0.05)
summary(is.de_tr)

# visualize results
plotSmear(tr, de.tags=rownames(tr)[is.de_tr!=0])
title(main="DE genes using glmTreat with logFC cutoff")
```

Be sure to knit this file into a pdf or html file once you're finished.

System information for reproducibility:

```{r}
pander::pander(sessionInfo())
```


