---
title: "Clustering"
author: "Carson Stacy & Jeffrey Lewis"
date: "Fall 2023"
output: 
  html_document:
    embed_resources: TRUE
---

# WGCNA

last updated: `r Sys.Date()`

```{r setup-wgcna, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
set.seed("1492")
```

**Install Packages**

As usual, make sure we have the right packages for this exercise

```{r ready-packages-wgcna, cache=FALSE}
if (!require("pacman")) install.packages("pacman"); library(pacman)

# let's load all of the files we were using and want to have again today
p_load("tidyverse", "knitr", "readr",
       "pander", "BiocManager", 
       "dplyr", "stringr", 
       "statmod", # required dependency, need to load manually on some macOS versions.
       "purrr", # for working with lists (beautify column names)
       "reactable") # for pretty tables.

# for today:
# p_load("WGCNA")   # WGCNA is available on CRAN
if (!require("WGCNA")) BiocManager::install("WGCNA", force=TRUE); library(WGCNA)


# 
# # We also need these Bioconductor packages today.
p_load("edgeR")
p_load("matrixStats")
p_load("org.Sc.sgd.db")
p_load("AnnotationDbi")
p_load("cowplot")
p_load("igraph")

```

### Description

Use WGCNA to identify modules identified in patterns of
transcriptomic changes across samples spanning multiple genotypes and stress conditions. Through this exercise, we aim to gain practical insights into interpreting WGCNA outputs and leveraging this technique for a deeper understanding of genomic landscapes and network analysis.

## Learning Objectives

At the end of this exercise, you should be able to:

-   Understand the concept of co-expression modules.
-   Interpret module colors and visualizations.
-   Evaluate module preservation and critically analyze results.


### Loading in the count data file

*We are downloading the expected raw counts from a
Github repository using the code below. Just as in previous exercises,
assign the data to the variable `counts`. You can change the file path
if you have saved it to your computer in a different location.*

```{r load-counts-raw-wgcna}
# Load in raw counts for all samples
counts <- read.delim('https://github.com/clstacy/GenomicDataAnalysis_Fa23/raw/main/data/YPS606_TF_Rep1_4_ExpectedCounts.txt',
    sep = "\t",
    header = T,
    row.names = 1 # convert first column to rownames
  )
```

### Preprocessing raw count data
```{r tidy-wgcna}
# tidy raw counts to visualize distributions
tidy_counts <- counts %>%
  rownames_to_column("gene_id") %>%
  tidyr::pivot_longer(
    .,                        # The dot is the the input data, magrittr tutorial
    col = -gene_id
    ) %>%  
  separate_wider_delim(name, delim = "_", names = c("strain", "stress", "replicate"),cols_remove = FALSE)

(
p <- tidy_counts %>%
    ggplot(., aes(x = replicate, y = value)) +        # x = treatment, y = RNA Seq count
    geom_violin() +                                   # violin plot, show distribution
    geom_point(alpha = 0.2) +                         # add transparent points
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 90)          # Rotate treatment text
    ) +
    labs(x = "Treatment Groups", y = "RNA Seq Counts") +
    facet_grid(cols = vars(stress), rows = vars(strain), drop = TRUE, scales = "free_x") +     # Facet by condition
    # scale_y_log10() +
    ggtitle("Raw count distributions")
)
```


### Data normalization

Let's use edgeR to normalize the raw counts
```{r normalize-edgeR-wgcna}
# read into a DGElist
y <- edgeR::DGEList(counts,
                    group=as.factor(
                      # remove replicate from name
                      sub("_[^_]+$", "", colnames(counts))
                    ),
                    genes = rownames(counts)
                    )

# filter low counts
keep <- rowSums(cpm(y) > 1) >= 4
y <- y[keep,]

# normalize with cpm
normalized_counts <- cpm(y,
                         prior.count = 4,
                         normalized.lib.sizes = FALSE,
                         log = TRUE)

# normalized_counts <- cpmByGroup(y,
#                          log = TRUE)
str(normalized_counts)
```

```{r}
pca <- prcomp(t(normalized_counts))

# Extract PC loadings and variances
pc_loadings <- pca$x
pc_variances <- pca$sdev^2

# Create PCA plot
pc_loadings %>%
  as.data.frame() %>%
  rownames_to_column("genotype_stress_replicate") %>%
  tidyr::separate(genotype_stress_replicate, into = c("genotype", "stress", "replicate"), sep = "_") %>%
ggplot(data = ., aes(x = PC1, y = PC2, color = stress)) +
  geom_point() +
  # geom_text(aes(label = rownames(pc_loadings)), nudge_x = 0.02, nudge_y = 0.02) +
  labs(
    title = "PCA plot of normalized gene expression",
    x = "PC1",
    y = "PC2"
  ) +
  facet_wrap(~genotype) +
  theme_bw()
```


Now that we've normalized, let's look at the distributions again

```{r tidy-normalized-data-wgcna}
# create data 
tidy_normalized_counts <- normalized_counts %>%
  data.frame() %>%
  rownames_to_column("gene_id") %>%
  tidyr::pivot_longer(
    .,                        # The dot is the the input data, magrittr tutorial
    col = -gene_id
    ) %>% # pull(name) %>% unique()
  tidyr::separate_wider_delim(name, delim = "_", names = c("strain", "stress", "replicate"),cols_remove = FALSE)

(
p_normalized <- tidy_normalized_counts %>%
    ggplot(., aes(x = replicate, y = value)) +        # x = treatment, y = RNA Seq count
    geom_violin() +                                   # violin plot, show distribution
    geom_point(alpha = 0.2) +                         # add transparent points
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 90)          # Rotate treatment text
    ) +
    labs(x = "Treatment Groups", y = "normalized expression") +
    facet_grid(cols = vars(stress),rows = vars(strain), drop = TRUE, scales = "free_x") +     # Facet by condition
    # scale_y_log10() +
    ggtitle("Normalized CPM distribution")
)
```

## Variability-based subsetting of gene list

For WGCNA, we don't want to correlate all of the genes because (1) computational limitation and (2) genes that vary a small amount increase the number of variables in the model, making it harder to be confident in the patterns we do see. One way to account for this is by subsetting our entire gene list. 

One common approach for subsetting genes is taking genes whose variance is greater than the $n^{th}$ quantile of gene varainces. We'll do that now:

```{r get-rowwise-variances-wgcna}
# get the variance for each row (gene)
rv_wpn <- matrixStats::rowVars(normalized_counts)

# see a summary of the quantiles of variances
summary(rv_wpn)
```




```{r find-quantile-rowwise-variances-wgcna}
# get variance corresponding to 75th percentile
q75_wpn <- quantile( matrixStats::rowVars(normalized_counts), .3)  # common cutoff

# # get variance corresponding to 95th percentile
# q95_wpn <- quantile( matrixStats::rowVars(normalized_counts), .95)  # <= changed to 95 quantile 

# to reduce dataset, only get genes with variance greater than above
normalized_counts_variable_genes <- normalized_counts[ rv_wpn > q75_wpn, ] %>% 
  t() %>%
  scale() %>% 
  t()

str(normalized_counts_variable_genes)
```
Notice the number of genes retained is much fewer now.

We can see the distribution of gene variances, and where our cutoff is
```{r}
ggplot(data=NULL,aes(x=rv_wpn)) +
  geom_histogram(bins=600) +
  geom_vline(xintercept = q75_wpn, color="red") +
  theme_bw() +
  ggtitle("Distribution of gene variances")
```

What are the distributions of counts of the normalized cpm's of the retained genes?
```{r see-distro-normalized-variable-genes-wgcna}
tidy_normalized_counts_variable_genes <- normalized_counts_variable_genes %>%
  data.frame() %>%
  rownames_to_column("gene_id") %>%
  pivot_longer(-gene_id) %>%  
  separate_wider_delim(name, delim = "_", names = c("strain", "stress", "replicate"),cols_remove = FALSE)

(
p_normalized_varGenes <- tidy_normalized_counts_variable_genes %>%
    ggplot(., aes(x = replicate, y = value)) +        # x = treatment, y = RNA Seq count
    geom_violin() +                                   # violin plot, show distribution
    geom_point(alpha = 0.2) +                         # add transparent points
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 90)          # Rotate treatment text
    ) +
    labs(x = "Treatment Groups", y = "normalized expression") +
    facet_grid(cols = vars(stress),rows = vars(strain), drop = TRUE, scales = "free_x") +     # Facet by condition
    # scale_y_log10() +
    ggtitle("Normalized CPM distribution")
)
```

For the analyses we've done so far, we've needed the columns to correspond to samples and the rows to correspond to genes. WGCNA expects the opposite, where columns are genes and rows are samples. We can use the transpose function in r `t()` in order to convert our data. Let's do that now:

```{r tranpose-variable-genes-wgcna}
# transpose our normalized count matrix and save to new object
variable_genes_matrix = t(normalized_counts_variable_genes)

# Look at first 5 rows and 10 columns
variable_genes_matrix[1:5,1:10]           
```

### WGCNA thresholding

We can see now that the `rows = treatments` and `columns = gene probes`. We’re ready to start WGCNA. A correlation network would be a complete network (all genes would be connected to all other genes). Hence we need to pick a threshhold value (if correlation is below threshold, remove the edge). We assume the true biological network follows a scale-free structure (see papers by Albert Barabasi).

To do that, WGCNA will try a range of soft thresholds and create a diagnostic plot. This step can take several minutes.

```{r}
# library(WGCNA)
allowWGCNAThreads(nThreads = 4)          # allow multi-threading (optional)
#> Allowing multi-threading with up to 4 threads.

# Choose a set of soft-thresholding powers
powers = c(c(1:10), seq(from = 12, to = 20, by = 2))

# Call the network topology analysis function
sft = pickSoftThreshold(
  variable_genes_matrix, # <= Input data
  #blockSize = 30,
  powerVector = powers,
  verbose = 5
  )
```
Let's visualize those results

```{r}
par(mfrow = c(1,2));
cex1 = 0.9;

plot(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     xlab = "Soft Threshold (power)",
     ylab = "Scale Free Topology Model Fit, signed R^2",
     main = paste("Scale independence")
)
text(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     labels = powers, cex = cex1, col = "red"
)
abline(h = 0.90, col = "red")
plot(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     xlab = "Soft Threshold (power)",
     ylab = "Mean Connectivity",
     type = "n",
     main = paste("Mean connectivity")
)
text(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     labels = powers,
     cex = cex1, col = "red")
```

Pick a soft threshold power near the curve of the plot. We’ll pick 8, but feel free to experiment with other powers to see how it affects your results. Now we can create the network using the `blockwiseModules` command. The `blockwiseModule` may take a while to run, since it is constructing the TOM (topological overlap matrix) and several other steps. While it runs, take a look at the `blockwiseModule` documentation (link to [vignette][https://www.rdocumentation.org/packages/WGCNA/versions/1.69/topics/blockwiseModules]) for more information on the parameters. How might you change the parameters to get more or less modules?

### Generate network

```{r}
# NOTE: Be sure to run this entire chunk at once
picked_power = 12
cor <- WGCNA::cor         # Force it to use WGCNA cor function (fix a namespace conflict issue)
variable_genes_network <- blockwiseModules(variable_genes_matrix,                # <= input here

                          # == Adjacency Function ==
                          power = picked_power,                # <= power here
                          networkType = "signed",

                          # == Tree and Block Options ==
                          deepSplit = 2,
                          pamRespectsDendro = F,
                          # detectCutHeight = 0.75,
                          minModuleSize = 30,
                          maxBlockSize = 8000,

                          # == Module Adjustments ==
                          reassignThreshold = 0,
                          mergeCutHeight = 0.25,

                          # == TOM == Archive the run results in TOM file (saves time)
                          saveTOMs = F,
                          saveTOMFileBase = "ER",

                          # == Output Options
                          numericLabels = F,
                          verbose = 3)

# return cor() to be the usual command
cor <- stats::cor
```

Let's visualize a dendrogram of those modules, each terminal branch corresponds to a single gene.
```{r fig.height=4, fig.width=8, dpi=600}
# Convert labels to colors for plotting
mergedColors_var_genes = #labels2colors(
  variable_genes_network$colors
  #)
# Plot the dendrogram and the module colors underneath
plotDendroAndColors(
  variable_genes_network$dendrograms[[1]],
  # mergedColors_var_genes[variable_genes_network$blockGenes],
  mergedColors_var_genes[variable_genes_network$blockGenes[[1]]],
  " ",
  dendroLabels = FALSE,
  hang = 0.03,
  addGuide = TRUE,
  guideHang = 0.05 )
```

Now let's extract the genes and their coresponding module/color to a data frame
```{r}
modules_variable_genes_df <- data.frame(
  gene_id = names(variable_genes_network$colors),
  colors = #labels2colors(
    variable_genes_network$colors
  # )
)

head(modules_variable_genes_df)

modules_variable_genes_df %>%
  group_by(colors) %>%
  summarise(n())
```

### Visualize Modules

Plot gene expression correlations in a grid, based on genotype and module membership
```{r fig.height=12, fig.width=8}
(
  var_genes_plot <- normalized_counts_variable_genes %>%
  data.frame() %>%
    mutate(
        gene_id = row.names(.)
    ) %>%
    pivot_longer(-gene_id) %>%
    mutate(
        module = variable_genes_network$colors[gene_id]
    ) %>% 
  separate_wider_delim(name, delim = "_", names = c("strain", "stress", "replicate"),cols_remove = FALSE) %>% 
    mutate(xlab = paste(stress, replicate, sep = "_")) %>%
  ggplot(., aes(x=xlab, y=value, group=gene_id)) +
    geom_line(aes(color = "grey"), linewidth=1,
            alpha = 0.2) +
    geom_line(aes(color = module),
            alpha = 0.2) +
    geom_boxplot(width=0.5,
                 aes(group=name, color = module),
                 outlier.shape = NA,
                 alpha=0.9) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 90)) +
    facet_grid(rows = vars(module), 
             cols= vars(fct_rev(as.factor(strain))),
             scales = "free"
             ) +
    scale_x_discrete(position = "top") +
    labs(x = "Environmental Condition",
       y = "Normalized Gene Expression (z-score)") +
    # use color variable as the color
    scale_color_identity()
)
```



### Heatmap correlating modules

Now, let's generate a heatmap of the relationship between the modules

```{r}
# Get Module Eigengenes per cluster
MEs0 <- moduleEigengenes(variable_genes_matrix, mergedColors_var_genes)$eigengenes

# Reorder modules so similar modules are next to each other
MEs0 <- orderMEs(MEs0)
module_order = names(MEs0) %>% gsub("ME","", .)

# Add treatment names
MEs0$treatment = row.names(MEs0)

# tidy & plot data
mME = MEs0 %>%
  pivot_longer(-treatment) %>%
  mutate(
    name = gsub("ME", "", name),
    name = factor(name, levels = module_order)
  )

mME %>% ggplot(., aes(x=treatment, y=name, fill=value)) +
  geom_tile() +
  theme_bw() +
  scale_fill_gradient2(
    low = "blue",
    high = "red",
    mid = "white",
    midpoint = 0,
    limit = c(-1,1)) +
  theme(axis.text.x = element_text(angle=90)) +
  labs(title = "Module-trait Relationships", y = "Modules", fill="corr")
```


### Generate & Visualize Networks

We can generate a network graph file for analysis in Cytoscape, or as an edge/vertices file.

```{r fig.height=12, fig.width=12, dpi=600}
# correlation cutoff
corr_cutoff = 0.3

# first, create adjacency matrix
adjacency = adjacency(variable_genes_matrix, power=picked_power, type="signed")

# Calculate topological overlap matrix and dissimilarity from adjacency matrix
TOM = TOMsimilarity(adjacency, TOMType="signed")

# Add gene names to row and columns
rownames(TOM) = AnnotationDbi::select(org.Sc.sgd.db,keys=row.names(adjacency),columns="GENENAME") %>% mutate(label = case_when(
  is.na(GENENAME) ~ ORF,
  TRUE ~ GENENAME)) %>% pull(label)


colnames(TOM) = AnnotationDbi::select(org.Sc.sgd.db,keys=row.names(adjacency),columns="GENENAME") %>% mutate(label = case_when(
  is.na(GENENAME) ~ ORF,
  TRUE ~ GENENAME)) %>% pull(label)

# assign TOM to new variable
adj <- TOM

# only show correlations greater than corr_cutoff
adj[adj > corr_cutoff] = 1

# set all correlations below that threshold to 0
adj[adj != 1] = 0

# graph adjacency network
network_graph_data <- graph.adjacency(adj)

# remove redundant self-loops in the network
network_graph_data <- simplify(network_graph_data)

# add colors for the network graph from constructed network modules
V(network_graph_data)$color <- variable_genes_network$colors

# set plotting margins
par(mar=c(0,0,1,0))
# remove unconnected nodes
network_graph_data <- delete.vertices(network_graph_data,
                                      degree(network_graph_data)==0)


# plot(network_graph_data, 
#      layout=layout.fruchterman.reingold(network_graph_data, niter=1000),
#      vertex.size=3,edge.arrow.size=0.1, 
#      vertex.label.cex = 0.01,  # increase this value to see gene names
#      vertex.label.color= "black", rescale=TRUE, ylim=c(-1,1),xlim=c(-1,1),
#      main = "Network plot for most variable genes")



# require("intergraph")
# require("ggnetwork")
# ggplot(ggnetwork::ggnetwork(network_graph_data,
#                      layout = with_fr()),
#                      aes(x = x, y = y, 
#                          xend = xend, yend = yend,
#                          label = name, color = color)) +
#   geom_edges(arrow = arrow(length = unit(1, "pt"), 
#                            type = "closed"),
#              # arrow.gap = 0,
#              color = "grey50", alpha=0.3) +
#   geom_nodes(size = 4.1, aes(color="black")) +
#   geom_nodes(size = 4) +
#   # geom_nodetext(check_overlap = TRUE,size=0.5) +
#   scale_color_identity() +
#   theme_blank() +
#   xlim(c(0,1)) +
#   ylim(c(0,1)) +
#   ggtitle("Network plot for most variable genes")
  
# plot(network_graph_data, 
#      layout=layout.fruchterman.reingold(network_graph_data, niter=1000),
#      vertex.size=3, edge.arrow.size=0.1, 
#      vertex.label.cex = 0.01,  # increase this value to see gene names
#      vertex.label.color= "black", rescale=TRUE, ylim=c(-1,1),xlim=c(-1,1),
#      main = "Network plot for most variable genes")
# 
# # Step 1: Call the pdf command to start the plot
# tiff(file = "~/Desktop/network_for_seminar.tiff",   # The directory you want to save the file in
#     width = 1200, # The width of the plot in inches
#     height = 1200, # The height of the plot in inches
#     res = 600)

# Step 2: Create the plot with R code

plot(network_graph_data, 
       layout=layout.fruchterman.reingold(network_graph_data, niter=1000),
       vertex.size=3, edge.arrow.size=0.1, 
       vertex.label.cex = 0.01,  # increase this value to see gene names
       vertex.label.color= "black", rescale=TRUE, ylim=c(-1,1),xlim=c(-1,1),
       main = " ")


# Step 3: Run dev.off() to create the file!
# dev.off()

```


### Save network for analysis in cytoscape
```{r}
# choose where to save files
dir_cytoscape_files <- path.expand("~/Desktop/Genomic_Data_Analysis/Analysis/Network/WGCNA")

#create that directory if it doesn't exist
if (!dir.exists(dir_cytoscape_files)) {dir.create(dir_cytoscape_files, recursive = TRUE)}

# save files for loading into Cytoscape
exportNetworkToCytoscape(
  adjMat = adjacency,
  edgeFile = paste0(dir_cytoscape_files,"variable_gene_network_edges.tsv"),
  nodeFile = paste0(dir_cytoscape_files,"variable_gene_network_nodes.tsv"),
  threshold=0.3,
  nodeAttr = modules_variable_genes_df$colors
)[["edgeData"]] %>% View()
```

### GO enrichments

Let's add GO enrichments to the module pattern graph, so we can make sense of what genes are composing each module.
```{r fig.height=18, fig.width=8}
# create list of genes in each one
variable_geneIDs_module_list <- normalized_counts_variable_genes %>%
  data.frame() %>%
    mutate(
        gene_id = row.names(.)
    ) %>%
    pivot_longer(-gene_id) %>%
    mutate(
        module = variable_genes_network$colors[gene_id]
    ) %>%
  group_by(module) %>%
  # only get genes that are DE
  # filter(DE_direction != 0) %>%
  group_split() %>% # split into many lists
  purrr::set_names(purrr::map_chr(., ~.x$module[1])) %>% # assign names
  # get just the ORF names
  map(., ~ .x |> pull(all_of("gene_id"))) %>%
  # remove duplicates
  map(., ~ unique(.))
  

# run enrichment for each module
GO_enrich_variable_gene_modules <- clusterProfiler::compareCluster(
                                  geneCluster   = variable_geneIDs_module_list, 
                                  ont           = "BP",
                                  OrgDb         = org.Sc.sgd.db,
                                  pAdjustMethod = "BH",
                                  pvalueCutoff  = 0.5, # high to get for all modules
                                  qvalueCutoff  = 0.5,
                                  readable      = FALSE,
                                  fun           = 'enrichGO',
                                  keyType       = 'ORF'
                                  )  |>
  # let's add a 'richFactor' column that gives us the proportion of genes DE in the term
  mutate(richFactor = Count / as.numeric(sub("/\\d+", "", BgRatio)))


# plot those enrichments
GO_enrich_variable_gene_modules %>%
  as.data.frame() %>% 
  group_by(Cluster) %>%
  slice_min(order_by=pvalue, n = 5) %>%
  ggplot(.,
           aes(Cluster, fct_reorder(Description, richFactor))) +
      geom_point(aes(color = p.adjust, size = Count)) +
      scale_color_gradientn(
        colours = c("#f7ca64", "#46bac2", "#7e62a3"),
        trans = "log10",
        guide = guide_colorbar(reverse = TRUE, order = 1)
      ) +
      scale_size_continuous(range = c(2, 12)) +
      scale_x_discrete(position = "top") +
      scale_y_discrete(labels = function(x) str_wrap(x, width = 25)) +
      xlab("Module") +
      ylab(NULL) +
      ggtitle("Enriched GO Categories by Module") +
      # facet_wrap(~DE, scales = "free_y") +
      theme_bw() +
      theme(text = element_text(size=14),
            axis.text.x = element_text(angle = 90,hjust=0) )
```

This next code chunk adds those GO terms alongside the module expression graph.
```{r fig.height=12, fig.width=8}
# here are the data to add to cowplot:
GO_summaries_variable_genes <- GO_enrich_variable_gene_modules %>%
  as.data.frame() %>% 
  group_by(Cluster) %>%
  slice_min(order_by=p.adjust, n = 5) %>%
  filter(nchar(Description) < 50) %>%
  slice_head(n = 5) %>%
  mutate(Description = case_when(
    p.adjust < 0.05 ~ paste0(Description," (n=",Count,")"),
    TRUE ~ " "
  )) %>%
  summarize(description = str_c(Description, collapse = "\n")) %>%
  pull(description)


# draw plot with GO terms
(
var_genes_plot_GO <- 
    cowplot::ggdraw(var_genes_plot +
                  theme(plot.margin = margin(t = 0.5, r = 5, b = 0.5, l = 0.1, unit = "cm"))) +
  cowplot::draw_text(GO_summaries_variable_genes, 
                     x = 0.76,  # Adjust this value for the desired horizontal position
                     y = rev(seq(0.1, 0.82, length.out = length(GO_summaries_variable_genes))),  # Adjust this sequence for the desired vertical positions
                     size = 6, 
                     hjust = 0,  # Align to the left
                     vjust = 0.5)
)

var_genes_plot_GO
```


## Create a network using genes identified as differentially expressed

Our network so far didn't use any prior information about genes to include or exclude, just those that are most variable. We do have information about the genes that are considered differentially expressed in their corresponding genotype and environment. What does the network look like if we build it based on those genes that have been identified as differentially expressed?

```{r}
# set FDR threshold for being considered a DE gene, 
# this is applied to the data we are loading in here.
FDR_cutoff <- 0.01

# Load in DE gene summary
DE_genes <- readr::read_delim("https://github.com/clstacy/GenomicDataAnalysis_Fa23/raw/main/data/DE_yeast_TF_stress.txt.gz",
    delim = "\t", 
    escape_double = FALSE, 
    name_repair = "universal",
    show_col_types=FALSE,
    trim_ws = TRUE) %>%
  dplyr::select(ID, contains("FDR")) %>%
  dplyr::select(ID, contains(".v.")) %>%
  pivot_longer(-ID, names_to = "contrast", values_to = "FDR") %>%
  # only get genes with FDR < 0.01
  filter(FDR < FDR_cutoff) %>%
  pull(ID) %>%
  unique()

# subset the counts to only be counts for genes that are DE
counts_DE <- counts %>%
  rownames_to_column("ORF") %>%
  dplyr::filter(ORF %in% DE_genes) %>%
  column_to_rownames("ORF")

# normalize with cpm, filter to just DE genes
normalized_counts_DE_genes <- cpm(y,
                         log = TRUE) %>%
  data.frame() %>%
  rownames_to_column("ORF") %>%
  dplyr::filter(ORF %in% DE_genes) %>%
  column_to_rownames("ORF") %>%
  as.matrix()


DE_genes_matrix = t(normalized_counts_DE_genes)
```

What is the correct power for this analysis?

### Find ideal power
```{r}
# Choose a set of soft-thresholding powers
powers = c(c(1:10), seq(from = 12, to = 20, by = 2))

# Call the network topology analysis function
sft = pickSoftThreshold(
  DE_genes_matrix,             # <= Input data
  #blockSize = 30,
  powerVector = powers,
  verbose = 5
  )

par(mfrow = c(1,2));
cex1 = 0.9;

plot(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     xlab = "Soft Threshold (power)",
     ylab = "Scale Free Topology Model Fit, signed R^2",
     main = paste("Scale independence")
)
text(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     labels = powers, cex = cex1, col = "red"
)
abline(h = 0.90, col = "red")
plot(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     xlab = "Soft Threshold (power)",
     ylab = "Mean Connectivity",
     type = "n",
     main = paste("Mean connectivity")
)
text(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     labels = powers,
     cex = cex1, col = "red")
```
It looks like a higher power would better fit this data. Let's try with 12. You can try changing to a different threshold and see how the results change.

```{r}
picked_power = 12
cor <- WGCNA::cor         # Force it to use WGCNA cor function (fix a namespace conflict issue)
DE_genes_network <- blockwiseModules(DE_genes_matrix,                # <= input here

                          # == Adjacency Function ==
                          power = picked_power,                # <= power here
                          networkType = "signed",

                          # == Tree and Block Options ==
                          deepSplit = 2,
                          pamRespectsDendro = F,
                          # detectCutHeight = 0.75,
                          minModuleSize = 30,
                          maxBlockSize = 4000,

                          # == Module Adjustments ==
                          reassignThreshold = 0,
                          mergeCutHeight = 0.25,

                          # == TOM == Archive the run results in TOM file (saves time)
                          saveTOMs = T,
                          saveTOMFileBase = "ER",

                          # == Output Options
                          numericLabels = F,
                          verbose = 3)

# fix namespace issue with cor to return to desired function
cor <- stats::cor
```

```{r fig.height=8, fig.width=12}
# Convert labels to colors for plotting
mergedColors_DE = labels2colors(DE_genes_network$colors)
# Plot the dendrogram and the module colors underneath
plotDendroAndColors(
  DE_genes_network$dendrograms[[1]],
  mergedColors_DE[DE_genes_network$blockGenes[[1]]],
  "Module colors",
  dendroLabels = FALSE,
  hang = 0.03,
  addGuide = TRUE,
  guideHang = 0.05 )
```


Now let's extract the genes and their corresponding module/color to a data frame
```{r}
modules_DE_genes_df <- data.frame(
  gene_id = names(variable_genes_network$colors),
  colors = labels2colors(variable_genes_network$colors)
)

head(modules_DE_genes_df)
```



### Visualize cluster patterns
```{r fig.height=12, fig.width=8}
(
  DE_genes_plot <- normalized_counts_DE_genes %>%
  data.frame() %>%
    mutate(
        gene_id = row.names(.)
    ) %>% 
    pivot_longer(-gene_id) %>%
    mutate(
        module = DE_genes_network$colors[gene_id]
    ) %>% 
  separate_wider_delim(name, delim = "_", names = c("strain", "stress", "replicate"),cols_remove = FALSE) %>%
  ggplot(., aes(x=name, y=value+1, group=gene_id)) +
    geom_line(aes(color = "grey"), size=1,
            alpha = 0.2) +
    geom_line(aes(color = module),
            alpha = 0.2) +
    # geom_boxplot(width=0.5,
    #              aes(group=name, color = module),
    #              outlier.shape = NA,
    #              alpha=0.9) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 90)) +
    facet_grid(rows = vars(module), 
             cols= vars(strain),
             scales = "free"
             ) +
    scale_x_discrete(position = "top") +
    labs(x = "treatment",
       y = "normalized expression") +
    # use color variable as the color
    scale_color_identity()
)

```



Generate a heatmap of the eigengenes
```{r}
# Get Module Eigengenes per cluster
DE_eigengenes <- moduleEigengenes(DE_genes_matrix, mergedColors_DE)$eigengenes

# Reorder modules so similar modules are next to each other
DE_eigengenes <- orderMEs(DE_eigengenes)

# get order of modules without extra naming
module_order = names(DE_eigengenes) %>% gsub("ME","", .)

# Add treatment names
DE_eigengenes$treatment = row.names(DE_eigengenes)

# tidy & plot data
tidy_DE_eigengenes <- DE_eigengenes %>%
  pivot_longer(-treatment) %>%
  mutate(
    name = gsub("ME", "", name),
    name = factor(name, levels = module_order)
  )

tidy_DE_eigengenes %>% 
  ggplot(., aes(x=treatment, y=name, fill=value)) +
  geom_tile() +
  theme_bw() +
  scale_fill_gradient2(
    low = "blue",
    high = "red",
    mid = "white",
    midpoint = 0,
    limit = c(-1,1)) +
  theme(axis.text.x = element_text(angle=90)) +
  labs(title = "Module-trait Relationships", y = "Modules", fill="corr")
```


### Generate and Export Networks

The network file can be generated for Cytoscape or as an edge/vertices file.

```{r}
genes_of_interest_DE = modules_DE_genes_df #%>%
  #subset(colors %in% modules_of_interest)

expr_of_interest_DE = normalized_counts_variable_genes#[genes_of_interest$gene_id,]


# Only recalculate TOM for modules of interest (faster, altho there's some online discussion if this will be slightly off)
TOM_DE = TOMsimilarityFromExpr(t(expr_of_interest_DE),
                            power = picked_power)

# Add gene names to row and columns
row.names(TOM_DE) = row.names(expr_of_interest_DE)
colnames(TOM_DE) = row.names(expr_of_interest_DE)

edge_list_DE = data.frame(TOM_DE) %>%
  mutate(
    gene1 = row.names(.)
  ) %>%
  pivot_longer(-gene1) %>%
  dplyr::rename(gene2 = name, correlation = value) %>%
  unique() %>%
  subset(!(gene1==gene2))

head(edge_list_DE)
```

What does the distribution of correlations look like?
```{r fig.height=5, fig.width=6}
edge_list_DE %>%
  ggplot(aes(x=correlation)) +
  geom_histogram(bins=60) +
  theme_bw()
```


Let's create the network based on these DE genes

```{r fig.height=12, fig.width=12}
# correlation cutoff
corr_cutoff = 0.3

# first, create adjacency matrix
adjacency_DE = adjacency(DE_genes_matrix, power=picked_power, type="signed")

# Calculate topological overlap matrix and dissimilarity from adjacency matrix
TOM_DE = TOMsimilarity(adjacency_DE, TOMType="signed")

# Add gene names to row and columns
rownames(TOM_DE) = AnnotationDbi::select(org.Sc.sgd.db,keys=row.names(adjacency_DE),columns="GENENAME") %>% mutate(label = case_when(
  is.na(GENENAME) ~ ORF,
  TRUE ~ GENENAME)) %>% pull(label)


colnames(TOM_DE) = AnnotationDbi::select(org.Sc.sgd.db,keys=row.names(adjacency_DE),columns="GENENAME") %>% mutate(label = case_when(
  is.na(GENENAME) ~ ORF,
  TRUE ~ GENENAME)) %>% pull(label)

# assign TOM to new variable
adj_DE <- TOM_DE

# only show correlations greater than corr_cutoff
adj_DE[adj_DE > corr_cutoff] = 1

# set all correlations below that threshold to 0
adj_DE[adj_DE != 1] = 0

# graph adjacency network
network_graph_data_DE <- graph.adjacency(adj_DE)

# remove redundant self-loops in the network
network_graph_data_DE <- simplify(network_graph_data_DE)

# add colors for the network graph from constructed network modules
V(network_graph_data_DE)$color <- DE_genes_network$colors


# remove unconnected nodes
network_graph_data_DE <- delete.vertices(network_graph_data_DE,
                                      degree(network_graph_data_DE)==0)

# set plotting margins
par(mar=c(0.5,0.5,0.5,0.5))

# generate plot
plot(network_graph_data_DE, 
     layout=layout.fruchterman.reingold(network_graph_data_DE, niter=1000),
     vertex.size=3, edge.arrow.size=0.1, 
     vertex.label.cex = 0.01,  # increase this value to see gene names
     vertex.label.color= "black", rescale=TRUE, ylim=c(-1,1),xlim=c(-1,1),
     main = "Network plot for differentially expressed genes")
```

### GO enrichments

Let's do GO enrichments on these terms as well.
```{r fig.height=18, fig.width=8}
# create list of genes in each one
DE_geneIDs_module_list <- normalized_counts_DE_genes %>%
  data.frame() %>%
    mutate(
        gene_id = row.names(.)
    ) %>%
    pivot_longer(-gene_id) %>%
    mutate(
        module = DE_genes_network$colors[gene_id]
    ) %>%
  group_by(module) %>%
  # only get genes that are DE
  # filter(DE_direction != 0) %>%
  group_split() %>% # split into many lists
  purrr::set_names(purrr::map_chr(., ~.x$module[1])) %>% # assign names
  # get just the ORF names
  map(., ~ .x |> pull(all_of("gene_id"))) %>%
  # remove duplicates
  map(., ~ unique(.))
  


GO_enrich_DE_gene_modules <- clusterProfiler::compareCluster(
                                  geneCluster   = DE_geneIDs_module_list, 
                                  ont           = "BP",
                                  OrgDb         = org.Sc.sgd.db,
                                  pAdjustMethod = "BH",
                                  pvalueCutoff  = 0.5, 
                                  qvalueCutoff  = 0.5,
                                  readable      = FALSE,
                                  fun           = 'enrichGO',
                                  keyType       = 'ORF'
                                  )  |>
  # let's add a 'richFactor' column that gives us the proportion of genes DE in the term
  mutate(richFactor = Count / as.numeric(sub("/\\d+", "", BgRatio)))


# get rid of sticks to save space
GO_enrich_DE_gene_modules %>%
  as.data.frame() %>% 
  group_by(Cluster) %>%
  slice_min(order_by=pvalue, n = 5) %>%
  ggplot(.,
           aes(Cluster, fct_reorder(Description, richFactor))) +
      geom_point(aes(color = p.adjust, size = Count)) +
      scale_color_gradientn(
        colours = c("#f7ca64", "#46bac2", "#7e62a3"),
        trans = "log10",
        guide = guide_colorbar(reverse = TRUE, order = 1)
      ) +
      scale_size_continuous(range = c(2, 12)) +
      scale_x_discrete(position = "top") +
      scale_y_discrete(labels = function(x) str_wrap(x, width = 25)) +
      xlab("Module") +
      ylab(NULL) +
      ggtitle("Enriched GO Categories by Module") +
      # facet_wrap(~DE, scales = "free_y") +
      theme_bw() +
      theme(text = element_text(size=14),
            axis.text.x = element_text(angle = 90,hjust=0) )
```


```{r fig.height=12, fig.width=8}
# here are the data to add to cowplot:
GO_summaries_DE_genes <- GO_enrich_DE_gene_modules %>%
  as.data.frame() %>% 
  group_by(Cluster) %>%
  slice_min(order_by=p.adjust, n = 5) %>%
  slice_head(n = 5) %>%
  mutate(Description = case_when(
    p.adjust < 0.05 ~ paste0(Description," (",Count,")"),
    TRUE ~ " "
  )) %>%
  summarize(description = str_c(Description, collapse = "\n")) %>%
  pull(description)


# draw plot with GO terms
cowplot::ggdraw(DE_genes_plot +
                  theme(plot.margin = margin(t = 0.5, r = 5, b = 0.5, l = 0.1, unit = "cm"))) +
  cowplot::draw_text(GO_summaries_DE_genes, 
                     x = 0.76,  # Adjust this value for the desired horizontal position
                     y = rev(seq(0.05, 0.82, length.out = length(GO_summaries_DE_genes))),  # Adjust this sequence for the desired vertical positions
                     size = 6, 
                     hjust = 0,  # Align GO terms to the left
                     vjust = 0.5)
```


## Saving network for Cytoscape

```{r}
# choose where to save files
dir_cytoscape_files <- path.expand("~/Desktop/Genomic_Data_Analysis/Analysis/Network/WGCNA")

#create that directory if it doesn't exist
if (!dir.exists(dir_cytoscape_files)) {dir.create(dir_cytoscape_files, recursive = TRUE)}

# save files for loading into Cytoscape
exportNetworkToCytoscape(
  adjMat = adjacency_DE,
  edgeFile = paste0(dir_cytoscape_files,"DE_gene_network_edges.tsv"),
  nodeFile = paste0(dir_cytoscape_files,"DE_gene_network_nodes.tsv"),
  threshold=0.3,
  nodeAttr = modules_DE_genes_df$colors
)
```
```

## Questions

Question 1: How would you explain the concept of a "module" in the context of WGCNA to someone unfamiliar with genomics?

Question 2: Why is the choice of soft power important in WGCNA, and how might it impact the results?

Question 3: How can the information obtained from WGCNA be integrated with other omics data for a more comprehensive biological interpretation?

Question 4: What does module preservation in WGCNA mean, and how to the two models generated with DE vs highly variable genes compare?

Be sure to knit this file into a pdf or html file once you're finished.

System information for reproducibility:

```{r sessionInfo-WGCNA}
pander::pander(sessionInfo())
```
